Direction-aware functional class scoring enrichment analysis of Infinium DNA methylation data

### 2024-02-14 manuscript/manuscript.Rmd
ms

DNA methylation is the most widely studied epigenetic mark, for its role in
development and disease [@Robertson2005-tu].
Hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
methylation patterns in disease and other contexts [@Li2019-og].
Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
as compared to high throughput methylation sequencing [@Wei2021-qj].
Infinium arrays typically include multiple probes per gene in different locations including promoters,
CpG islands, gene bodies and enhancers [@Pidsley2016-ix], which complicates downstream functional enrichment analysis,
otherwise known as pathway analysis or gene set enrichment analysis.
Enrichment analysis comes in two popular types: over-representation analysis (ORA) and functional
class scoring (FCS) [@Khatri2012-qy].
A third type of enrichment analysis called pathway topology improves upon other methods with more sophisticated
modelling of gene network activations, but these are yet to be adopted widely [@Xie2021-py].
In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
to a background list of all genes measured in the assay.
The test seeks to identify sets of genes (e.g., ontologies, pathways) that are over-represented in the
gene list of interest relative to the background [@Tavazoie1999-ng].
FCS takes a different approach by ranking all detected genes by a differential
regulation score (e.g., fold change, confident effect size, t-statistic) followed by a test to assess whether each
set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of infinium methylation array
data as it addresses issues related to probes belonging to more than one gene and the fact that one gene can have multiple
probes [@Maksimovic2021-ur].
As gsameth is an ORA method, results strongly depend on the significance threshold used [@Ren2019-mr], and also
on the proportion of probes that meet this threshold.
FCS methods `ebGSEA` and `methylGSA` have been developed for EWAS data and are suggested to have better sensitivity to
determine subtle associations between pathways and differential methylation [@Dong2019-np;@Ren2019-mr].
The ebGSEA tool uses an empirical Bayes modeling approach followed by a one-tailed nonparametric test for gene set
enrichment [@Dong2019-np].
The `methylGSA` package implements two different approaches.
Firstly, the `methylRRA` function, which uses robust rank aggregation for detecting differentially methylated genes taking
into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside each gene set
[@Ren2019-mr].
These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
direction of methylation change, it limits the utility for downstream interpretation of the importance of methylation
changes on genomic regulation.
Moreover, as pathways tend to be either upregulated or downregulated in omics assays, combining the two tends to dilute
their signal and reduce sensitivity [@Hong2014-ab].
Neither gsameth nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
enrichment score is a useful surrogate for effect size.
Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
biases towards large gene sets that have non-specific biological functions.

Here we aim to develop and evaluate methods for two-tailed FCS of Infinium methylation array data that addresses these
limitations.
Since genes have a variable number of probes, there are many concevable ways that FCS could be applied
to Infinium methylation array data.
The first set of methods (type I) use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
The differential methylation of probes beloning to each gene are scored and these gene scores
undergo an enrichment test.
The second set of methods (type II) first aggregate probe methylation values for each gene before differential
methylation analysis with limma.
Then, differential gene methylation results are used for a downstream enrichment test.
With the above methods there are several methodological choices, such as the type of aggregation, and the type of
downstream test.
To determine the best approach to applying FCS to Infinium methylation data, in this study, we use
simulated data and compare results to existing ORA methods.
We then examine the sensitivity of selected methods on real cancer data, investigate the association with gene expression
and end with additional examples related to human ageing, assisted reproductive technologies and a large-scale EWAS of
19 diseases.

@@

4. Limma Average Mitch (LAM).
Similar to LAW, however instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
two-sample ANOVA-on-ranks test as described previously[@Cox2012-zu].

5. Limma Rank Mitch (LRM).
Probes are ranked by t-statistic. The mean rank values for probes beloning to each gene are computed and used
as input for a two-way, two-sample ANOVA-on-ranks test.

6. Aggregate Limma t-test (ALT).
In this method, all probe measurements beloning to each gene are averaged before conducting differential
methylation andlysis with limma.
The limma t-statistics are then used downstream for enrichment analysis with the two-sample two-way t-test.

7. Aggregate Limma Wilcox (ALW).
Similar to ALT, however it uses the Wilcoxon non-parametric test for gene set enrichment.

8. Aggregate Limma Mitch (ALM).
Similar to ALW, however uses the Mitch package to execute a two-way, two-sample ANOVA-on-ranks test.

@@

To assess these methods, we adopted a simulated data approach based on the selected modification of real methylation data.
We downloaded raw intensity EPIC IDAT files from NCBI Gene Expression Omnibus (GEO) for study GSE158422, which consists of
lung tumour and normal adjacent tissues from 37 patients [@Cho2022-yd].
For the simulations, only non-cancerous datasets were used.
Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
package.
We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
3 and 12.
One thousand random gene sets were created with sizes varying between 20 and 100 members, with member genes drawn from the
annotation set.
Random gene sets were used to avoid problems caused by the large overlap among real gene sets.
Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
with decreased methylation.
From these gene sets, half of the member genes were selected.
For those selected genes, half of the annotated probes were selected.
We adjusted the M-values in the case group by a specified amount, which we call the "delta", which we varied
between 0.1 and 0.5.
Following the incorporation of methylation changes to selected probes in the case group, the data underwent
limma differential methylation analysis, followed by enrichment analysis using `gsameth`, or one of
the eight FCS methods outlined above.
Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
At each setting for group size, delta and gene set size, 100 replications were conducted, each with a set seed
to ensure reproducibility.
Simulations were conducted on a SLURM cluster [@Yoo2003-mc].
Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.

@@

RNA-seq based gene counts for the same set of lung cancer patients were downloaded from GEO (accession: GSE158420).
HGNChelper was used to fix gene names converted to dates, followed by DESeq2 based differential analysis.
This underwent filtering to remove genes with expression below 10 reads per sample on average across the dataset.
Then DESeq2[@Love2014-wt] was used to compare normal and cancerous tissue gene expression taking into consideration sample pairing.
The differential expression results underwent enrichment analysis using the mitch package [@Kaspi2020-yx] with default settings for
DESeq2 data tables.
CpG sites annotated as promoters were considered separately from those located at gene bodies in mitch analysis.
For comparison, GSA was used to separately examine pathways using separate analyses of promoter and gene body CpGs.
For this analysis, Reactome gene sets were used.
Subsequently, to generate the pathway heatmap, we used mitch in multivariate mode by analysing differential RNA expression,
promoter and gene body methylation together, prioritising the results by s-distance, a surrogate measure of effect size.

@@

In order to deminstrate the utility of the LAM method for large EWAS studies, we examined
EPIC methylation profile associations with age of two independent cohorts involving a total of 7036
participants [@McCartney2020-ob].
Limma summary statistics for chronological age including discovery and replication groups were read into R and enrichment analysis was conducted with the LAM method.
Briefly, gene-probe associations were first obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
dataset, followed by update of gene symbols with HGNChelper.
Discovery and replication data were imported separately with mitch and multivariate enrichment analysis was conducted with gene sets from Reactome.

@@

In order to demonstrate that methylation patterns across different array platforms can be compared using the LAM method,
we examined methylation differences in infants associated with assisted reproduction as described by two
independent studies.
The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO Accession GSE79257
[@Estill2016-jh].
The Novakovic study in was published 2019, and used the EPIC methylation array and the data is available from GEO under accession
GSE131433[@Novakovic2019-jt].
Various assisted reproductive technology conception groups are described in these studies, but we only examined
fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalistion was conducted with the SWAN method
[@Maksimovic2012-hy], and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes
located on X or Y chromosomes.
M values were computed and underwent differential methylation analysis using limma [@Ritchie2015-zn].
The Estill study included 43 naturally conceived infants and 38 with fresh embryo transfer.
The Novakovic study included 58 naturally conceived infants and 75 with fresh embryo transfer.
Limma differential analysis was conducted accounting for sex, then methylation tables underwent LAM analysis with Reactome gene sets.
Probe gene associations for HM450K array were established using the "IlluminaHumanMethylation450kanno.ilmn12.hg19" annotation set
and the gene symbols were updated with HGNChelper.
As the Novakovic study involved EPIC array data, gene-probe associations were obtained from the
"IlluminaHumanMethylationEPICanno.ilm10b4.hg19" annotation set, and gene symbols were updated with
HGNChelper.

@@

To examine whether LAM can identify differentially methylated pathways associated with disease prevalence and incidence,
we downloaded blood based EWAS summary statistics from a recent publication [@Hillary2023-mu], made available via
Zenodo (https://zenodo.org/records/8021411).
The study used a logitudinal design, with sample collection baseline and diagnosis status established via electronic health
record linkage up to 14 years after baseline.
The study included 18,413 volunteers of European ancestry.
We used summary statistics from the "full model", which corrected for relevant potential confounders.
Disease status at baseline was used to classify participants into prevalence groups.
Disease status at follow-up was used to classify participants into incidence groups.
The study used the EPIC Infinium array and 752,722 probes were described in the summary statistic dataset.
As limma t-statistics were not available, we used probe delta-beta values as input for LAM.
LAM was conducted individually for each disease condition with Reactome gene sets.
Finally, as a test of type I errors, we randomised each of the 19 incident profiles using
base R `sample` command with a set seed, followed by LAM enrichment and counting the number
of significant pathways.
This process was repeated 100 times with unique seeds.

@@

Associations of epigenetic marks with gene expression are of great interest for understanding disease processes.
Matching tumour-normal RNA-seq datasets were analysed and we conducted enrichment analysis of gene expression together with
DNA methylation.
From 18,704 detected genes, there were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated
in the tumour group respectively.
At the pathway level, there were 304 and 135 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
As the context of gene methylation is important in influencing gene expression, promoter and gene body methylation
were considered separately.
In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were were
18 and 75 pathways with higher and lower methylation respectively (FDR<0.05).
Using LAM at promoters, there were 190 and 20 pathways identified with higher and lower methylation respectively.
LAM at gene bodies identified 251 and 31 pathways with higher and lower methylation respectively (FDR<0.05).
The relative overlap between significant promoter and gene body pathways was larger for LAM pathway sets (22%) as compared to GSA
(3%) (**Figure 5A,B**).
There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).

@@

We sought to explore the utility of LAM for exploring large EWAS datasets.
One of the best studied associations with DNA methylation is with chronological aging, although
pathway level differential methylation is not well defined.
Using limma summary statistics from independent discovery and replication studies using the EPIC array platform described
by McCartney et al (2020) [@McCartney2020-ob], we applied the LAM method to explore the pathway level
differential methylation.
In the discovery cohort we observed 304 differential pathways (FDR<0.05), while in the replication group we observed 107.
There was a high degree of agreement, with 43 pathways significant in both groups.
This is not surprising, as the gene-level rank differential methylation scores show a strong positive association between discovery and
replication studies (**Figure 6A**).
Using the rank-MANOVA test of the mitch package, we identified 390 pathways with altered methylation.
Visualised as a scatterplot, the overall pattern is concordant, although a cluster of points in the lower right of the chart
indicates 268 pathways with higher methylation in the discovery sample and lower in methylation in the replication
sample (**Figure 6B**).
When prioritising results by S distance, the enrichment score reported by mitch and a surrogate measure of effect size,
29 of the top 30 pathways were concordant in their direction of regulation (**Figure 6C**).
Some of the observed methylation changes make sense with reports in the literature.
For example, the complement pathway is associated with aging and in this analysis we observe a reduction in methylation of genes
beloning to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).

@@

At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with a prevous report of mitch's performance with
simulated RNA-seq gene expression data[@Kaspi2020-yx].
The improved sensitivity of the LAM method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
subtle signatures, as we have demonstrated with the chronological age, assisted reproduction examples and 19 disease EWAS analyses.
The analyses conducted here used Reactome pathways, but LAM is general purpose, so different gene sets can be used to examine various hypotheses.
Transcription factor target gene sets can be used to identify transcription factors associated with changes in pathway methylation.
MicroRNA target gene sets can identify potential associations between microRNA targets and DNA methylation patterns.
The MSigDB resource contains these and several other types of gene sets for exploration [@Liberzon2015-um].

@@

The analysis of the large-scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
In contrast, LAM identified 2,469 pathway associations (74.8 per model).
To demonstrate these are not the results of false positives, permutated profiles yielded on average 0.10 significant findings per model,
supporting the idea that LAM is identifying real pathway signatures.

@@

There is also potential to generate new libraries of differentially methylated probe and gene sets, to contribute to the pool of molecular signatures in public resources like MSigDB,
which will assist in understanding similarities between methylation profiles.

>>>>

DNA methylation is the most widely studied epigenetic mark, for its role in
development and disease [@Robertson2005-tu].
Hundreds of epigenome-wide association studies (EWASs) are conducted each year to understand DNA
methylation patterns in disease and other contexts [@Li2019-og].
Infinium arrays remain the preferred platform for EWASs due to low cost and analytical simplicity
as compared to high throughput methylation sequencing [@Wei2021-qj].
Infinium arrays typically include multiple probes per gene in different locations including promoters,
CpG islands, gene bodies and enhancers [@Pidsley2016-ix], which complicates downstream functional enrichment analysis,
otherwise known as pathway analysis or gene set enrichment analysis.
Enrichment analysis comes in two popular types: over-representation analysis (ORA) and functional
class scoring (FCS) [@Khatri2012-qy].
A third type of enrichment analysis called pathway topology improves upon other methods with more sophisticated
modelling of gene network activations, but these are yet to be adopted widely [@Xie2021-py].
In ORA, genes with probes that meet an arbitrary significance threshold are selected and compared
to a background list of all genes measured in the assay.
The test seeks to identify sets of genes (e.g., ontologies, pathways) that are over-represented in the
gene list of interest relative to the background [@Tavazoie1999-ng].
FCS takes a different approach by ranking all detected genes by a differential
regulation score (e.g., fold change, confident effect size, t-statistic) followed by a test to assess whether each
set of genes has a distribution of scores that is different to the null [@Subramanian2005-ar].
The `gsameth()` function of the `missMethyl` package is the state-of-the-art method for ORA of Infinium methylation array
data as it addresses issues related to probes belonging to more than one gene and the fact that one gene can have multiple
probes [@Maksimovic2021-ur].
As gsameth is an ORA method, results strongly depend on the significance threshold used [@Ren2019-mr], and also
on the proportion of probes that meet this threshold.
FCS methods `ebGSEA` and `methylGSA` have been developed for EWAS data and are suggested to have better sensitivity to
determine subtle associations between pathways and differential methylation [@Dong2019-np;@Ren2019-mr].
The ebGSEA tool uses an empirical Bayes modeling approach followed by a one-tailed nonparametric test for gene set
enrichment [@Dong2019-np].
The `methylGSA` package implements two different approaches.
Firstly, the `methylRRA` function, which uses robust rank aggregation for detecting differentially methylated genes taking
into consideration the variable number of probes followed by an FCS test based on z-scores [@Ren2019-mr].
Secondly, `methylglm` uses a logistic regression approach to model differential methylation for genes inside and outside each gene set
[@Ren2019-mr].
These methods are indeed more sensitive than ORA approaches, but as they are based on one-tailed tests irrespective of
direction of methylation change, it limits the utility for downstream interpretation of the importance of methylation
changes on genomic regulation.
Moreover, as pathways tend to be either upregulated or downregulated in omics assays, combining the two tends to dilute
their signal and reduce sensitivity [@Hong2014-ab].
Neither gsameth nor methylGSA report computed enrichment scores, which hinders downstream interpretation, as the
enrichment score is a useful surrogate for effect size.
Without an enrichment score, users are limited to using statistical significance values to prioritise results, which
biases towards large gene sets that have non-specific biological functions.

Here we aim to develop and evaluate methods for two-tailed FCS of Infinium methylation array data that addresses these
limitations.
Since genes have a variable number of probes, there are many conceivable ways that FCS could be applied
to Infinium methylation array data.
The first set of methods (type I) use differential probe methylation data from `limma` [@Ritchie2015-zn] as an input.
The differential methylation of probes belonging to each gene are scored and these gene scores
undergo an enrichment test.
The second set of methods (type II) first aggregate probe methylation values for each gene before differential
methylation analysis with limma.
Then, differential gene methylation results are used for a downstream enrichment test.
With the above methods there are several methodological choices, such as the type of aggregation, and the type of
downstream test.
To determine the best approach to applying FCS to Infinium methylation data, in this study, we use
simulated data and compare results to existing ORA methods.
We then examine the sensitivity of selected methods on real cancer data, investigate the association with gene expression
and end with additional examples related to human ageing, assisted reproductive technologies and a large-scale EWAS of
19 diseases.

@@

4. Limma Average Mitch (LAM).
Similar to LAW, however, instead of the Wilcoxon test, the mitch package[@Kaspi2020-yx] is used to execute a two-way,
two-sample ANOVA-on-ranks test as described previously[@Cox2012-zu].

5. Limma Rank Mitch (LRM).
Probes are ranked by t-statistic. The mean rank values for probes belonging to each gene are computed and used
as input for a two-way, two-sample ANOVA-on-ranks test.

6. Aggregate Limma t-test (ALT).
In this method, all probe measurements belonging to each gene are averaged before conducting differential
methylation analysis with limma.
The limma t-statistics are then used downstream for enrichment analysis with the two-sample two-way t-test.

7. Aggregate Limma Wilcox (ALW).
Similar to ALT, however, it uses the Wilcoxon non-parametric test for gene set enrichment.

8. Aggregate Limma Mitch (ALM).
Similar to ALW, however, uses the Mitch package to execute a two-way, two-sample ANOVA-on-ranks test.

@@

To assess these methods, we adopted a simulated data approach based on the selected modification of real methylation data.
We downloaded raw intensity EPIC IDAT files from NCBI Gene Expression Omnibus (GEO) for study GSE158422, which consists of
lung tumour and normal adjacent tissues from 37 patients [@Cho2022-yd].
For the simulations, only non-cancerous datasets were used.
Probe annotations were obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" Bioconductor
package.
We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
3 and 12.
One thousand random gene sets were created with sizes varying between 20 and 100 members, with member genes drawn from the
annotation set.
Random gene sets were used to avoid problems caused by the large overlap among real gene sets.
Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
with decreased methylation.
From these gene sets, half of the member genes were selected.
For those selected genes, half of the annotated probes were selected.
We adjusted the M-values in the case group by a specified amount, which we call the "delta", which we varied
between 0.1 and 0.5.
Following the incorporation of methylation changes to selected probes in the case group, the data underwent
limma differential methylation analysis, followed by enrichment analysis using `gsameth`, or one of
the eight FCS methods outlined above.
Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
At each setting for group size, delta and gene set size, 100 replications were conducted, each with a set seed
to ensure reproducibility.
Simulations were conducted on a SLURM high performance computing cluster [@Yoo2003-mc].
Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.

@@

RNA-seq based gene counts for the same set of lung cancer patients were downloaded from GEO (accession: GSE158420).
HGNChelper was used to fix gene names converted to dates, followed by DESeq2 based differential analysis.
This underwent filtering to remove genes with expression below 10 reads per sample on average across the dataset.
Then DESeq2[@Love2014-wt] was used to compare normal and cancerous tissue gene expression taking into consideration sample pairing.
The differential expression results underwent enrichment analysis using the mitch package [@Kaspi2020-yx] with default settings for
DESeq2 data tables.
CpG sites annotated as promoters were considered separately from those located at gene bodies in mitch analysis.
For comparison, GSA was used to separately examine pathways using separate analyses of promoter and gene body CpGs.
For this analysis, Reactome gene sets were used.
Subsequently, to generate the pathway heatmap, we used mitch in multivariate mode by analysing differential RNA expression,
promoter and gene body methylation together, prioritising the results by S distance, a surrogate measure of effect size.

@@

In order to demonstrate the utility of the LAM method for large EWAS studies, we examined
EPIC methylation profile associations with age of two independent cohorts involving a total of 7036
participants [@McCartney2020-ob].
Limma summary statistics for chronological age including discovery and replication groups were read into R and enrichment analysis was conducted with the LAM method.
Briefly, gene-probe associations were first obtained from the "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
dataset, followed by update of gene symbols with HGNChelper.
Discovery and replication data were imported separately with mitch and multivariate enrichment analysis was conducted with gene sets from Reactome.

@@

In order to demonstrate that methylation patterns across different array platforms can be compared using the LAM method,
we examined methylation differences in infants associated with assisted reproduction as described by two
independent studies.
The Estill study was published in 2016 and is based on the HM450K array and the data is available from GEO Accession GSE79257
[@Estill2016-jh].
The Novakovic study in was published 2019, and used the EPIC methylation array and the data is available from GEO under accession
GSE131433[@Novakovic2019-jt].
Various assisted reproductive technology conception groups are described in these studies, but we only examined
fresh in vitro fertilisation (IVF) conceived infants compared to naturally conceived infants.
Raw intensity files (IDAT format) were read into R with minfi [@Aryee2014-su], normalisation was conducted with the SWAN method
[@Maksimovic2012-hy], and probe filtering was conducted to remove proves with a detection p-value > 0.01 in addition to probes
located on X or Y chromosomes.
M values were computed and underwent differential methylation analysis using limma [@Ritchie2015-zn].
The Estill study included 43 naturally conceived infants and 38 with fresh embryo transfer.
The Novakovic study included 58 naturally conceived infants and 75 with fresh embryo transfer.
Limma differential analysis was conducted accounting for sex, then methylation tables underwent LAM analysis with Reactome gene sets.
Probe gene associations for HM450K array were established using the "IlluminaHumanMethylation450kanno.ilmn12.hg19" annotation set
and the gene symbols were updated with HGNChelper.
As the Novakovic study involved EPIC array data, gene-probe associations were obtained from the
"IlluminaHumanMethylationEPICanno.ilm10b4.hg19" annotation set, and gene symbols were updated with
HGNChelper.

@@

To examine whether LAM can identify differentially methylated pathways associated with disease prevalence and incidence,
we downloaded blood based EWAS summary statistics from a recent publication [@Hillary2023-mu], made available via
Zenodo (https://zenodo.org/records/8021411).
The study used a longitudinal design, with sample collection baseline and diagnosis status established via electronic health
record linkage up to 14 years after baseline.
The study included 18,413 volunteers of European ancestry.
We used summary statistics from the "full model", which corrected for relevant potential confounders.
Disease status at baseline was used to classify participants into prevalence groups.
Disease status at follow-up was used to classify participants into incidence groups.
The study used the EPIC Infinium array and 752,722 probes were described in the summary statistic dataset.
As limma t-statistics were not available, we used probe delta-beta values as input for LAM.
LAM was conducted individually for each disease condition with Reactome gene sets.
Finally, as a test of type I errors, we randomised each of the 19 incident profiles using
base R `sample` command with a set seed, followed by LAM enrichment and counting the number
of significant pathways.
This process was repeated 100 times with unique seeds.

@@

Associations of epigenetic marks with gene expression are of great interest for understanding disease processes.
Matching tumour-normal RNA-seq datasets were analysed and we conducted enrichment analysis of gene expression together with
DNA methylation.
From 18,704 detected genes, there were 12,380 differentially expressed genes (FDR<0.05), with 7,415 and 4,965 up- and down-regulated
in the tumour group respectively.
At the pathway level, there were 304 and 135 up- and down-regulated Reactome pathways with altered gene expression (FDR<0.05).
As the context of gene methylation is important in influencing gene expression, promoter and gene body methylation
were considered separately.
In promoters, GSA identified 15 and 7 pathways with higher and lower methylation respectively, while at gene bodies, there were
18 and 75 pathways with higher and lower methylation respectively (FDR<0.05).
Using LAM at promoters, there were 190 and 20 pathways identified with higher and lower methylation respectively.
LAM at gene bodies identified 251 and 31 pathways with higher and lower methylation respectively (FDR<0.05).
The relative overlap between significant promoter and gene body pathways was larger for LAM pathway sets (22%) as compared to GSA
(3%) (**Figure 5A,B**).
There was no observed overlap between GSA methylation-based pathways and gene expression pathways (**Figure 5A**), but there were some
overlaps between pathways identified with LAM based promoter and gene body methylation and gene expression (**Figure 5B**).

@@

We sought to explore the utility of LAM for exploring large EWAS datasets.
One of the best studied associations with DNA methylation is with chronological aging, although
pathway level differential methylation is not well defined.
Using limma summary statistics from independent discovery and replication studies using the EPIC array platform described
by McCartney et al (2020) [@McCartney2020-ob], we applied the LAM method to explore the pathway level
differential methylation.
In the discovery cohort we observed 304 differential pathways (FDR<0.05), while in the replication group we observed 107.
There was a high degree of agreement, with 43 pathways significant in both groups.
This is not surprising, as the gene-level rank differential methylation scores show a strong positive association between discovery and
replication studies (**Figure 6A**).
Using the rank-MANOVA test of the mitch package, we identified 390 pathways with altered methylation.
Visualised as a scatterplot, the overall pattern is concordant, although a cluster of points in the lower right of the chart
indicates 268 pathways with higher methylation in the discovery sample and lower in methylation in the replication
sample (**Figure 6B**).
When prioritising results by S distance, the enrichment score reported by mitch and a surrogate measure of effect size,
29 of the top 30 pathways were concordant in their direction of regulation (**Figure 6C**).
Some of the observed methylation changes make sense with reports in the literature.
For example, the complement pathway is associated with aging [@Gaya_da_Costa2018-sy]
and in this analysis we observe a reduction in methylation of genes
belonging to this set including *C1R*, *MBL2*, *C1S* and *C1QC* (**Figure 6D**).

@@

At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with a previous report of mitch's performance with
simulated RNA-seq gene expression data[@Kaspi2020-yx].
The improved sensitivity of the LAM method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
subtle signatures, as we have demonstrated with the chronological age, assisted reproduction examples and 19 disease EWAS analyses.
The analyses conducted here used Reactome pathways, but LAM is general purpose, so different gene sets can be used to examine various hypotheses.
Transcription factor target gene sets can be used to identify transcription factors associated with changes in pathway methylation.
MicroRNA target gene sets can identify potential associations between microRNA targets and DNA methylation patterns.
The MSigDB resource contains these and several other types of gene sets for exploration [@Liberzon2015-um].

@@

The analysis of the large-scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
In contrast, LAM identified 2,469 pathway associations (74.8 per model).
To demonstrate these are not the results of false positives, permuted profiles yielded on average 0.10 significant findings per model,
supporting the idea that LAM is identifying real pathway signatures.

@@

There is also potential to generate new libraries of differentially methylated probes and gene sets, to contribute to the pool of molecular signatures in public resources like MSigDB,
which will assist in understanding similarities between methylation profiles.
